---
title: 'Final Project: R Reference'
author: "Hazel"
date: "12/2/2021"
output: html_document
---

```{r}

```

# R Reference Guide {.tabset .tabset-pills}


## Loading Data and Packages

### library() and require()
The function library() and require() are both used to load and attach installed packages to R. R does not always load the extra packages you installed, so you need to use one of these functions to tell R to load the package. 

If the package is already loaded:

 * The library() function will re-load it.
 * The require() function will check first and will not re-load if it is already loaded.
 
I prefer to use the require() function because some packages take a long time to load, so not re-loading a package saves a lot of time. 

```{r}
## Using the library() function to load penguin data package
library(palmerpenguins)

## Using the require() function to load here package
require(here)
```


### here() and read.csv()

The function here() locates project files based on the current working directory. Using the function here() will return the absolute path to the base directory of your set RProject.

The read.csv() function loads csv files from the file system and stores the data in a data frame

 * Using the here() package is an easy way to access files and can access files from a different working directory without changing the main Rproject directory. 
 * While using here(), use subdirectories (in the correct order) and filename in quotation marks to assemble an absolute path to the file. 

The following is an example of how to use the here() and read.csv() functions:
```{r}
## Using 2021 ginkgo data to create a data frame using read.csv()

ginkgo = read.csv(here("data", "ginkgo_data_2021.csv"))

ginkgo
```

```{r}
## Using the here() function will show us the location of our data 

here("data", "ginkgo_data_2021.csv")
```


## Data Structures


### c()

The function c() combines or concatenates its arguments into a vector (a 1-dimensional data structure consisting of 1 or more elements).

 * All of the elements must be of the same type.
 * I can’t combine character and numeric types in the same call to c()

Here’s two examples using numeric and character data types:

```{r}
## Create a vector of numbers:
num_vec  = c(1, 4, 8, 9, 13)

## Create a vector of characters:
char_vec = c("a", "fish", "data is cool")
```

I can show the contents of a vector by typing the name of the vector, or using the print() function.

```{r}
## Typing the name of the vector into the console prints the contents
num_vec

## [1]  1  4  8  9 13

## The print() function accomplishes the same task:
print(char_vec)

## [1] "a"            "fish"         "data is cool"
```


### length()
The length() function determines how many elements are in a vector.

```{r}
## Create a vector of numbers from the previous c() example
num_vec  = c(1, 4, 8, 9, 13)


## The length function will give an output of the number of values in our example vector

length(num_vec)

```


### matrix()
The matrix() function constructs a matrix in R. This function has three arguments that lets the function know how the elements need to be arranged.

 * The first argument is the collection of elements that will be arranged by the function. You can use a vector as the first element.
 
 * The argument byrow will indicate if we want the matrix to be filled by rows. If we do, we place byrow = TRUE in the function, if we want the matrix filled with just columns, we place byrow = FALSE in the matrix() function.
 
 * The third argument, nrow, indicates how many rows the matrix should have and ncol indicates the columns.
 
 * Something to note while looking at the second example, I only created one row because the number of rows you include should be a multiple of your data length.
 
```{r}
## I will use the num_vec vector from the previous example to create a matrix that has five rows and four columns  
num_vec  = c(1, 4, 8, 9, 13)

num_matrix <- matrix(num_vec, nrow= 5, ncol= 3)

num_matrix

## I can also create a matrix with the num_vec vector that is filled row-wise rather than

num_matrix_2 <- matrix(num_vec, byrow= TRUE, nrow= 1)

num_matrix_2

```

### data.frame()

The data.frame() function is used to construct a data frame.A data frame can contain combination of different vectors, matrices, and arguments.
 
 * A data frame can contain different types of data such as numerical, logical, and character data.
 
 * The columns in a data frame each contain one value from the variables and each row contains one set of values from each columns.
 
 * All elements in a column have the same data type but different columns can have different data types.
 
```{r}
## Using the data.frame function to join two vectors into a new data frame

my_vec = rep(1:3, 5)

my_bool_vec<-my_vec == 3

data.frame(my_vec, my_bool_vec)

```

### Functions used with data.frame() and matrix()

nrow()
 * The nrow() returns the number of rows of a data frame or a matrix.
 
ncol()
 * The ncol() returns the number of columns of a data frame or a matrix.
 
dim()
 * The dim() function returns or assigns the number of rows and colums of a data frame or a matrix.
 
```{r}
## Using the ginkgo data frame to create examples of nrow(), ncol(), and dim()

ginkgo = read.csv(here("data", "ginkgo_data_2021.csv"))

nrow(ginkgo)

ncol(ginkgo)

dim(ginkgo)


```

## Subsetting

### subset()

Subsetting a data frame can be done by using a couple different functions, depending on what we want to accomplish.
 * To subset a data frame by name, we can use '$'.
 * To subset a data frame by position, we can use '[]'.
 * The subset() function can be used to extra a specific part of the input data based on given parameters.


```{r}
# First load the data frame that will be used

ginkgo = read.csv(here("data", "ginkgo_data_2021.csv"))


# Selecting the 'petiole_length' column in the gingko data frame

ginkgo$petiole_length

```

```{r}
# Selecting the first row of the ginkgo data

ginkgo[1,0]
```

```{r}
# Selecting the element in row 2, column 3 of the ginkgo data

ginkgo[2,3]

```

```{r}
# Selecting the third column of the ginkgo data

ginkgo[0,3]

```
```{r}
# Using the subset() function to retrieve all of the data for Adelie penguins from the species column in the penguins dataset

library(palmerpenguins)

penguin_ade <- subset(penguins, species == "Adelie")

```


## Numerical Data Exploration

There are several functions that can be used to calculate and analyze the numeric values of a data set. 

 * summary() is used to summarize the values in a vector, data frame, or other model functions.

 * mean() is used to calculate the sum of all values and divide it with the total number of values in a given data set.
 
 * sd() is used to quantify the amount of variation or dispersion of a set of given data values.


```{r}
# Using the ginkgo data to summarize the data frame

ginkgo = read.csv(here("data", "ginkgo_data_2021.csv"))


summary(ginkgo)

```

```{r}
# Using the mean() function to calculate the mean of the 'petiole_length' column in the ginkgo data frame 


mean(ginkgo$petiole_length, na.rm = TRUE)

# The na.rm() function indicates whether 'NA' values should be removed before computation  
```

```{r}
# Using the sd() function to calculate the standard deviation of the 'petiole_length' column in the ginkgo data frame

sd(gingko$petiole_length, na.rm = TRUE)
```
## Graphical Data Exploration

The plot() argument creates a scatterplot from given values/parameters. 

The hist() argument creates a histogram from given values/parameters.

The boxplot() argument creates a boxplot of given data from given values/parameters.

```{r}
# Using the ginkgo data to create a scatterplot of max leef depth (x) and max leaf width(y)

ginkgo = read.csv(here("data", "ginkgo_data_2021.csv"))

plot(x = gingko$max_depth, y = gingko$max_width, 
     xlab = "Max depth", ylab = "Max width", main = "Gingko leaves")

```

There are also several other arguments that can be used when making graphs that can help to create a better result. The following can all be used in a plot() graph:
 * col = Used to specify colors with the associated number in R. For example, the color red is denoted by col = 2. This is useful when plotting multiple groups because you can change each group to a different color.
 * pch = Used to change the dimensions of points on a plot graph. For example, pch = 1 is a standard circle, but can be changed to pch = 2 which is a triangle.
 * cex = Used to change the size of the plotting symbol. The default is 1, 1.5 is 50% larger, 0.5 is 50%, etc. 
 * main = Used to create a title. Example: main = "Title of Graph"
 * xlab = Used to label the x-axis of a graph. Example: xlab = "My x axis".
 * ylab = Used to label the y-axis of a graph. Example: ylab = "This is my y axis"
 * xlim = Used to set limits of the x-axis. For example, to create a scatterplot with x-axis limits ranging from 0-20, include xlim = c(0,20) in the plot() command. 
 * ylim = Used to set limits for the y-axis. For example, to create a scatterplot with y-axis limits ranging from 20-80, include ylim = c(20,80) in the plot() command. 

```{r}
# Create a scatterplot with the above plot() commands

ginkgo = read.csv(here("data", "ginkgo_data_2021.csv"))

plot(x = ginkgo$max_depth, y = ginkgo$max_width, 
     xlab = "Max depth", ylab = "Max width", main = "Ginkgo leaves", col = "2", pch = 24, cex = 1.5, xlim = c(0, max(ginkgo$max_depth)+20), ylim = c(0, max(ginkgo$max_width)+20))


```

```{r}
# Creating a histogram using the penguin flipper lengths of the palmer penguins data set.

# Using the breaks argument in a histogram will specify the number of cells we want.

library(palmerpenguins)

hist(penguins$flipper_length_mm, xlab = "Flipper length", main = "Penguin Flipper Length", breaks = 4)


```

```{r}
# Creating boxplots using the ginkgo data

ginkgo = read.csv(here("data", "ginkgo_data_2021.csv"))

# A simple boxplot of petiole lengths

boxplot(ginkgo$petiole_length, main = "Ginkgo Petiole Length")

# A conditional boxplot of one of the continuous variables conditioned on the seeds_present column

boxplot(data = ginkgo, max_width ~ seeds_present, "Max width by Seeds present", xlab = "Seeds present", ylab = "Max Leaf Width")

```

We can also make a figure out of the graphs that we make by using creating an output file by using the png() and dev.off().

 * par(mfrow = c()) is used within the png() function to arrange the rows and columns within the output figure. 

```{r}
# Using the ginkgo data to create a panel of four histograms

ginkgo = read.csv(here("data", "ginkgo_data_2021.csv"))

png(filename = "ginkgo_histograms.png", width = 800, height = 600)
par(mfrow = c(2,2))

hist(ginkgo$max_width, xlab = "Max Width", main = "Ginkgo Leaf Width", breaks = 9)
hist(ginkgo$petiole_length, xlab = "Petiole Length", main = "Ginkgo Leaf Petiole Length", breaks = 8)
hist(ginkgo$max_depth, xlab = "Max Depth", main = "Ginkgo Leaf Depth", breaks = 9)
hist(ginkgo$notch_depth, xlab = "Notch Depth", main = "Ginkgo Leaf Notch Depth", breaks = 6)


dev.off()
```

## Distribution Functions

The normal distribution functions are: dnorm, pnorm, and qnorm

dnorm(): Calculates the probability distribution of a specific event for a given mean and standard deviation. 
 * The mean and standard deviation must be provided.

```{r}
# Example of dnorm using a previous example

dnorm(x,mean,sd)

dnorm(1.5, mean = 0, sd = 1)
```

pnorm(): Returns the cumulative distribution for a given value in a normal distribution. 
 * The mean and standard deviation of the distribution must be provided.

```{r}
# Example of pnorm  

pnorm(q,mean,sd)

pnorm(1.5, mean = 0, sd = 1)

```

qnorm(): Returns the quantile for a given value in a normal distribution.
 * The mean and standard deviation must be provided. 
 
```{r}
# Example of qnorm

qnorm(p,mean,sd)
qnorm(1.5,mean = 0, sd = 1)

```

The binomial distribution functions are: dbinom, pbinom, and qbinom. 

dbinom(): the probability mass function, .
 * The size and probability must be provided. 
```{r}
# Example:
dbinom(x, size, prob)

```

pbinom(): the cumulative mass function, 


```{r}
pbinom(x, size, prob)
```

qbinom(): the quantile function

```{r}
qbinom(p, size, prob)

```
