---
title: "Final Data Analysis"
author: "Hazel Ortiz"
date: "12/7/2021"
output: html_document
---

I worked on this assignment with: Justin Maynard, Justin Roch, Aliza Fassler, Sonja Glasser

```{r}
require(here)

delomys = read.csv(here("data", "delomys.csv"))

```

## Data Exploration

### Numerical Exploration
```{r}
summary(delomys$body_mass)

summary(delomys$body_length)

```

```{r}
shapiro.test(delomys$body_mass)

shapiro.test(delomys$body_length)
```

### Graphical Exploration
```{r}
plot(x = delomys$body_mass, y = delomys$body_length, main = " Relationship of Delomys Body Length and Body Mass", xlab = "body length (?)", ylab = "body mass (g)")

```

```{r}
hist(delomys$body_mass, main = "Delomys Body Mass", xlab = "body mass (g)", breaks = 10)

```
```{r}
hist(delomys$body_length, main = "Delomys Body Length", xlab = "body length (?)")

```

```{r}
boxplot(body_mass ~ binomial, data = delomys, main = "Delomys Body Length for Species", xlab = "Species", ylab = "body mass (g)")
```

```{r}
boxplot(body_mass ~ sex, data = delomys, xlab = "Sex", ylab = "body mass (g)")
```

```{r}
boxplot(body_mass ~ sex + binomial, data = delomys, xlab = "Species by Sex", ylab = "body mass (g)")

```

#### Q1: 

The relationship between the body mass and length appears linear. There are a few outliers, and it almost appears there is a very slight curve associated with the data

#### Q2:

These two histograms are not normally distributed. They are skewed in one direction with a larger representation of data towards the left. Body mass seems to be more evenly distributed than length, although still not normally distributed.

#### Q3:

The p-values from the Shapiro test are very small, and the histogram is skewed in one direction. The p-values reject the null hypothesis, meaning that the data is not normally distributed since the Shapiro test null hypothesis is “The data are drawn from a normally-distributed population”. This is supported visually by the histograms, which are skewed.

#### Q4:

By reviewing the boxplot conditioned on species, the body mass appears to be larger with the dorsalis species. Looking at the boxplot conditioned by only sex, it appears that males are slightly larger than females. When conditioned by both variables, there are more patterns being developed regarding their variance and these changes are more reflective in the boxplot. When conditioned by both species and sex, males are larger in both species and the dorsalis species is larger over all than the sublineatus species.



## Model Diagnostics
```{r}

fit1 = lm(delomys$body_length ~ delomys$body_mass, data = delomys)
fit2 = lm(delomys$body_mass ~ delomys$sex, data = delomys)
fit3 = lm(delomys$body_mass ~ delomys$binomial, data = delomys)
fit4 = lm(delomys$body_mass ~ delomys$sex + delomys$binomial, data = delomys)
fit5 = lm(delomys$body_mass ~ delomys$sex * delomys$binomial, data = delomys)

```

## Residuals
```{r}

residuals(fit1)
residuals(fit2)
residuals(fit3)
residuals(fit4)
residuals(fit5)
```


### Histogram Model Residuals
```{r}
hist(residuals(fit1), main = "Model Residuals for Fit1", xlab = "Fit1 Residuals")

hist(residuals(fit2), main = "Model Residuals for Fit2", xlab = "Fit2 Residuals")

hist(residuals(fit3), main = "Model Residuals for Fit3", xlab = "Fit3 Residuals")

hist(residuals(fit4), main = "Model Residuals for Fit4", xlab = "Fit4 Residuals")

hist(residuals(fit5), main = "Model Residuals for Fit5", xlab = "Fit5 Residuals")
```


### Shapiro Test Model Residuals
```{r}
shapiro.test(residuals(fit1))
shapiro.test(residuals(fit2))
shapiro.test(residuals(fit3))
shapiro.test(residuals(fit4))
shapiro.test(residuals(fit5))
```

#### Q5:

The numerical data (Shapiro test) show that the models are not normally distributed but the histograms (except for fit1) visually appear they are normally distributed. 

#### Q6:

No. Fit 2 is the least severe violation because it has the largest p-value when looking at the Shapiro test.

## Model Interpretation
```{r}
summary(fit1)
summary(fit2)
summary(fit3)
summary(fit4)
summary(fit5)

```

## Model Coefficients 
```{r}
knitr::kable(coef(summary(fit1)))
knitr::kable(coef(summary(fit2)))
knitr::kable(coef(summary(fit3)))
knitr::kable(coef(summary(fit4)))
knitr::kable(coef(summary(fit5)))
```

#### Q7
The magnitude of the mass/length relationship is 0.8754988. With every unit increase of length (intercept), mass goes up 0.8754988.

#### Q8
The expected body length of an animal that weighs 100g is 163.624 grams. Using the equation y = mx + b and replacing values m with the magnitude .875, x with 100, and b with the intercept 76.124

Equation for 100 grams becomes y = (0.875)100 + 76.124 = 163.624

#### Q9
The expected body length of an animal weighing 0g is 76.12. 

Y = mx + b
Y = (.875)x + 76.124 = .875 x 0 + 76.124 = 76.124

#### Q10 
The base case is female (42.7115)

#### Q11
The base level for the binomial is Demolys dorsalis (46.7524)

#### Q12
Males are heavier (48.8658) 

#### Q13
Demolys dorsalis is heavier.

## Body Mass: ANOVA
```{r}
knitr::kable(anova(fit1))
knitr::kable(anova(fit2))
knitr::kable(anova(fit3))
knitr::kable(anova(fit4))
knitr::kable(anova(fit5))

```

#### Q14
Sex and species are both significant predictors for body mass.

#### Q15
There is not a significant interaction between sex and species.

#### Q16 
The p values for the main effect sex is slightly higher in the coefficient table than in the additive and interactive tables.  The p-values for the main effect in species is close to 0 in all models, so there is little effect. The p-value for the interaction of species and sex is 0.95. 

## Model Comparison: Body Mass 
```{r}
AIC(fit2)
AIC(fit3)
AIC(fit4)
AIC(fit5)
```

#### Q17 
Fit4 (body_mass ~ sex + binomial) and Fit5 (body_mass ~ sex * binomial ) both have the lowest AIC value, with fit4 being the lowest. 

#### Q18
I would choose fit 4 because it is an additive model and simpler than the factorial model. Looking at the AIC, fit 4 is slightly smaller making it the more optimal choice as well. The AIC measures goodness of fit and also penalizes for the number of parameters in a model, with the smaller AIC being optimal.
